---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

First, load the R packages for this analysis.

```{r, echo=TRUE, message=FALSE,results="hide"}
library(plyr)
library(dplyr)
library(lattice)
```

## Loading and preprocessing the data 
```{r, echo=TRUE}
data <- read.csv("./repdata-data-activity/activity.csv")
str(data)
```

## What is mean total number of steps taken per day? 
Missing values are ignored for this part.

**1. Calculate the total number of steps taken per day**
```{r, echo=TRUE}
data2 <- data[!is.na(data$steps),]
str(data2)
by_day <- group_by(data2,date)
sum <- summarise(by_day,totalsteps=sum(steps))
sum
```

**2. Make a histogram of the total number of steps taken each day**
```{r, echo=TRUE}
hist(sum$totalsteps, xlab="Total steps each day", main="Total number of steps taken each day")
```

**3. Calculate and report the mean and median of the total number of steps taken per day**
```{r, echo=TRUE}
mean(sum$totalsteps)
```
```{r, echo=TRUE}
median(sum$totalsteps)
```


## What is the average daily activity pattern? 
Missing values are ignored for this part.

**1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**
```{r, echo=TRUE}
by_interval <- group_by(data2,interval)
sum2 <- summarize(by_interval, avgstep = mean(steps))
sum2
plot(sum2$interval,sum2$avgstep, type="l", xlab="Interval", ylab="Number of steps")
```

**2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**
```{r, echo=TRUE}
maxstep <- max(sum2$avgstep)
sum2[sum2$avgstep==maxstep,]
```


## Imputing missing values

**1. The total number of missing values in the dataset**
```{r, echo=TRUE}
my_na <- is.na(data)
sum(my_na)
```

**2. Impute the missing value, using the mean for that 5-mintue interval**

Create a data frame to include NA only and merge it with summarized data set by interval to get mean for the 5-minute interval.
```{r, echo=TRUE}
x <- data[is.na(data),]
y <- left_join(x,sum2, by="interval")
# remove original column with NA values and use mean instead
z <- select(y,avgstep,date,interval)
colnames(z) <- c("steps","date","interval")
head(z,10)
```

**3. Create a new dataset that is equal to the original dataset but with the missing data filled in.**
```{r, echo=TRUE}
w <- rbind(data2,z)
w <- arrange(w,date,interval)
head(w)
summary(w)
```

**4. Make a histogram of the total number of steps taken each day**
```{r, echo=TRUE}
by_day_new <- group_by(w,date)
sum_new <- summarize(by_day_new,totalsteps=sum(steps))
sum_new
hist(sum_new$totalsteps, xlab="Total steps each day", main="Total number of steps taken each day")
```

**5. Calculate and report the mean and median total number of steps taken per day**
```{r, echo=TRUE}
mean(sum_new$totalsteps)
median(sum_new$totalsteps)
```
These values differ from the estimates from the first part of the assignment. After imputing the missing values, the median equals to the mean.


## Are there differences in activity patterns between weekdays and weekends?
**1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day**
```{r, echo=TRUE}
w$weekd <- weekdays(as.Date(as.character(w$date)))
w1 <- w[w$weekd==c("Sunday","Saturday"),]
w1$dayind <- "weekend"
w2 <- w[w$weekd!=c("Sunday","Saturday"),]
w2$dayind <- "weekday"
combinedw <- rbind(w1,w2)
head(combinedw)
# tail(combinedw)
```

**2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).**
```{r, echo=TRUE}
by_interval_new <- group_by(combinedw,dayind,interval)
sum2_new <- summarize(by_interval_new, avgstep = mean(steps))
# sum2_new
xyplot(avgstep ~ interval | dayind, data = sum2_new, layout = c(1,2), type = "l", xlab="Interval", ylab="Number of steps")
```
